Scaffolder - Software for Microbial Genome Scaffolding.

### 2011-02-25 src/article.tex
Add Hazel corrections to introduction

High throughput sequencing can produce hundreds of thousands to millions of
sequence reads from a genome. Each sequence represents the machine-read
nucleotides from a short region in the genome. At the time of writing, high
throughput sequencing is limited to producing reads less than 1000 nucleotides
in length. Therefore to determine a sequence longer than this, such as
a complete genome, these numerous individual reads must be pieced together. The
process of joining reads into longer sequences is the `assembly' stage in
a genome project \cite{miller2010}. \pb

Assembly software takes the nucleotide reads produced by the sequencing
hardware and, in the ideal case, outputs a complete genome sequence composed of
these individual reads. An analogy for this process is a jigsaw puzzle. Each
nucleotide read represents a single piece and the final genome sequence is the
completed puzzle. Sequences of repetitive nucleotide `repeat' regions within
the genome, or biased and incomplete sequencing data may however prevent the
genome being assembled into a continuous sequence. This is analogous to missing
jigsaw pieces or pieces that fit to multiple other pieces. \pb

The recent advent of these current high throughput sequencing methods has lead
to a renewed interest into algorithms for solving the problem of genome
assembly \cite{pop2008,pop2009}. Genome assembly software may however be unable
to produce a complete sequence from these reads. In many cases assembly
algorithms will generate several large contiguous regions of sequence
(`contigs') composed of many individual reads. These contigs represent the
assembly of reads into the longest possible sequences given the available data
and assembly algorithm. These contigs represent a fragmented picture of the
genome and require additional work to join together into larger sequence
lengths. \pb

The process of ameliorating a draft sequence into a finished continuous
sequence can be expensive in both time and laboratory effort. Therefore the
genomic data present in a set of contigs may often be enough for many research
questions \cite{branscomb2002}. A continuous high-quality `finished' genome
sequence does however provide a greater depth of information, such as complete
resolution of repeat regions and precise estimates of distances between genomic
elements \cite{parkhill2002,fraser2002}. This process of joining contigs
together to form a continuous sequence may be called the `scaffolding' or
`finishing' stage. \pb

@@

Scaffolding is the process of joining a series disconnected contigs into
complete continuous sequence. Due to genomic complexity and missing data,
scaffolding may not ultimately produce a final completed sequence, but may
still succeed in joining a subset of contigs together or filling gaps. An
overview of the scaffolding process is outlined below.

@@

The sequencing process generates reads from either strand of the DNA helix.
Therefore the resulting contigs constructed from these reads may point in
either direction of the DNA. The orientation of contigs to point in the same
direction and represent the same strand requires reverse complementing
sequences where necessary. In the case of microbes this orientation may be to
the 5'-3' direction leading away from the origin of replication.

@@

Contig ordering determines the placement of observed contigs to best represent
their order in the true genome sequence. Correctly placing each contig also
highlights any extra-genomic DNA, such as plasmids which should be scaffolded
separately from the genomic sequence. Contig order is commonly started at the
origin of replication where the first contig is split at the replication start
point. All subsequent contigs are then ordered in the direction of DNA
replication.

@@

Given the correct order and orientation, determining the distances between
contigs results in an estimate of the complete genome size. The size of inter
contig gaps represents the length of an unknown region in the genome. Filling
these regions with unknown nucleotide characters (`N') allows a draft
continuous sequence. This draft scaffold sequence is useful for representing
both the sequenced and unknown regions in the genome.

@@

During the scaffolding process closing and filling gaps between contigs
improves the completeness of the genome scaffold. Closing gaps may require
returning to the laboratory to perform additional sequencing or using
computational methods of estimating the unknown sequence. This additional
sequence is used to replace the gap between two contigs joining them into
a single sequence if they are not already scaffolded together. Once all contigs
have been joined and gaps in a scaffold closed the genome may be considered
finished.

@@

The process of resolving a genome sequence and building a scaffold uses
\emph{in vitro} methods, \emph{in silico} methods, or a combination of both. An
example of a computational method might use paired-read data, where the
sequencing reads are produced in pairs at a known distance apart in the genome.
The occurrence of a pair of reads in separate contigs can be used to estimate
the order and distance between the contigs. Alternatively laboratory methods
may use PCR to amplify the unknown DNA in a gap region, then use traditional
Sanger sequencing to determine the sequence. Computational methods are more
preferable as they are less costly in laboratory time and materials compared to
manual gap resolution \cite{nagarajan2010}. This need for computational
scaffolding methods has resulted in a variety of algorithmic approaches. \pb

Synteny methods compare the assembled contigs to a more complete reference
genome sequence searching for areas of sequence similarity between the two.
Areas of corresponding sequence between the the two sets are can be used to
infer contig placement and build the scaffold \cite{richter2007,zhao2008}.
Recombination between the two genomes can however reduce the effectiveness of
this analysis. \pb

Repeat regions may be responsible for many of the gaps in building a genome
sequence. The existence of tandemly repeated nucleotide regions produces many
similar reads in the sequence process. The assembly of reads generated from
repeat regions can result in the assemble collapsing into a single repeat
instance or ignored by conservative assembly algorithms. These regions may be
resolved by using reassembly algorithms specifically for repeat resolution
\cite{mulyukov2002,koren2010}. Similarly a related approach may use uniquely
matching reads to try and bridge across gaps between contigs \cite{tsai2010}.
\pb

As mentioned above, the genome may be prepared prior to sequencing so that
reads are produced in pairs at a known distance. This paired-read data
provides an extra level of information on how contigs may be scaffolded
together. Heuristic graph algorithms may take advantage of this data to search
for optimal configuration of contigs in the scaffold using paired-read distances
\cite{dayarian2010}. Additional data using a reference genome may also be
combined when estimating the best contig configuration \cite{pop2004}. Finally
when the scaffold cannot be completely resolved \emph{in silico} software
packages exist to suggest the likely primers necessary for PCR amplifying the
sequence in gap regions \cite{gordon2001,nagarajan2010}. \pb

\subsection*{Summary} %{{{2

Taken together these methods provide a wide variety of possible approaches for
scaffolding a genome. The scaffold may still however require manual editing and
addition of sequence if computational methods are not completely successful.
This involves moving and joining larges blocks of nucleotide text together by
hand which may be error-prone and precludes any reproducibility.
\pb

>>>>

High throughput sequencing can produce hundreds of thousands to millions of
sequence reads from a genome. Each sequence represents the machine-read
nucleotides from a short region in the genome. At the time of writing, high
throughput sequencing is limited to producing reads less than 1000 nucleotides
in length. Therefore to resolve a sequence longer than this, such as a complete
genome, these numerous smaller fragments must be pieced together. The process
of joining reads into longer sequences is the `assembly' stage of a genome
project \cite{miller2010}. \pb

Assembly software takes the nucleotide reads produced by the sequencing
hardware and, in the ideal case, outputs a complete genome sequence composed of
these individual fragments. An analogy for this process is a jigsaw puzzle.
Each nucleotide read represents a single piece and the final genome sequence is
the completed puzzle. Sequences of repetitive nucleotide `repeat' regions
within the genome, or biased and incomplete sequencing data, may prevent the
genome being assembled into a continuous sequence. This may be due to not
enough or multiple overlaps between sequence reads and is analogous to missing
pieces in the jigsaw or pieces that fit to multiple other pieces. \pb

The recent advent of these current high throughput sequencing methods has lead
to a renewed interest into algorithms for solving the problem of genome
assembly \cite{pop2008,pop2009}; genome assembly software may however be unable
to produce a complete sequence from these reads. In many cases assembly
algorithms will generate several large contiguous regions of sequence
(`contigs') composed of the many available individual reads, which represent the
assembly of reads into the longest possible sequences given the data.
Nonetheless, these contigs represent a fragmented picture of the genome and
require additional work to join together into larger sequence lengths. \pb

The process of ameliorating a draft sequence into a finished continuous
sequence can be expensive in both time and laboratory effort, although the
genomic data present in a set of contigs may often be sufficient for many
research questions \cite{branscomb2002}. Nevertheless, a continuous
high-quality `finished' genome sequence does provide a greater depth of
information, such as complete resolution of repeat regions and precise
estimates of distances between genomic elements \cite{parkhill2002,fraser2002}.
This process of joining contigs together to form a continuous sequence is
called the `scaffolding' or `finishing' stage. \pb

@@

Scaffolding is the process of joining a series disconnected contigs into
complete continuous genome sequence. Due to genomic complexity and missing
data, scaffolding may not ultimately produce a final completed sequence, but
may still succeed in joining a greater subset of contigs together or resolving
gaps. An overview of required steps in the scaffolding process is outlined
below.

@@

The sequencing process generates reads from either strand of the DNA helix, the
resulting contigs constructed from these reads may point in either direction of
the DNA. Orientating contigs to point in the same direction and represent the
same strand requires producing the reverse complement of sequences where
necessary. In the case of microbial genomes this orientation may be to the
5'-3' direction leading away from the origin of replication.

@@

Contig ordering determines the placement of observed contigs to best represent
their order in the true genome sequence. The Correct placement of each contig
also highlights any extra-genomic DNA, such as plasmids which are scaffolded
separately from the genomic sequence. The contig order is commonly started at
the contig containing the origin of replication. All subsequent contigs are
then ordered in the 5' - 3' direction of DNA replication.

@@

Given the correct order and orientation, determining the distance between
contigs results in an estimate of the complete genome size. The size of
inter-contig gaps represents the length of an unknown region in the genome.
Filling these regions with unknown nucleotide characters (`N') allows a draft
continuous sequence, which is useful for representing both the sequenced and
still to be resolved area in the genome sequence.

@@

During the scaffolding process closing and filling gaps between contigs
improves the completeness of the genome scaffold. Closing gaps may require
returning to the laboratory to perform additional sequencing or using
computational methods of estimating the unknown sequence. This additional
sequence is used to replace the gap between two contigs joining them into
a single sequence if they are not already scaffolded together. Once all contigs
have been joined and gaps in a scaffold closed, the genome may be considered
finished.

@@

The process of resolving a genome sequence and building a scaffold uses
\emph{in vitro} methods, \emph{in silico} methods, or a combination of both. An
example of a computational method might use paired-read data, where the
sequencing reads are produced in pairs at a known distance apart in the genome.
The occurrence of a pair of reads in separate contigs can be used to estimate
the order and distance between the contigs. Alternatively, laboratory methods
may use PCR to amplify the unknown DNA in a gap region, then use traditional
Sanger sequencing to determine the gap sequence. Computational methods are more
preferable as they are less costly in laboratory time and materials compared to
manual gap resolution and uses the available data that exists following
sequencing \cite{nagarajan2010}. \pb

Synteny methods compare the assembled contigs to a more complete reference
genome sequence searching for areas of sequence similarity between the two.
Areas of corresponding sequence between the two sets are can be used to infer
contig placement and build the scaffold \cite{richter2007,zhao2008,assefa2009}.
Recombination between the two genomes can however reduce the effectiveness of
this analysis. Repeat regions may be responsible for many of the gaps in
building a genome sequence. The existence of tandemly repeated nucleotide
regions produces many similar reads in the sequence process. The assembly of
reads generated from repeat regions can result in the assemble collapsing into
a single repeat instance or ignored by conservative assembly algorithms. These
regions may be resolved by using reassembly algorithms specifically for repeat
resolution \cite{mulyukov2002,koren2010}. Similarly a related approach may use
unassembled sequence reads matching the regions around a scaffold gap to
construct a uniquely overlapping set of reads across the region
\cite{tsai2010}. \pb

Paired-read data provides an extra level of information on how contigs may be
scaffolded together. Heuristic algorithms may take advantage of this data to
search for optimal configuration of contigs in the scaffold using paired-read
distances \cite{dayarian2010,boetzer2011}. Additional data using a reference
genome may also be combined when estimating the best contig configuration
\cite{pop2004}. Finally when the scaffold cannot be completely resolved
\emph{in silico} software packages exist to suggest the likely primers
necessary for PCR amplifying the sequence in gap regions
\cite{gordon2001,nagarajan2010}. \pb

These methods provide a wide variety of possible approaches for scaffolding
a genome. The scaffold may still however require manual editing and addition of
sequence if computational methods are not completely successful. This involves
moving and joining larges blocks of nucleotide text together by hand which may
be error-prone and precludes any reproducibility. \pb

\subsection*{Overview} %{{{2
